\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage[numbers]{natbib}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{xspace}

\newcommand{\At}{\ensuremath{\hat{A}_t}\xspace}

\title{Re-feeding Is Not Replaying:\\
Transcript Re-feeds Flip the Sign of\\
Counterfactual Token-Credit Estimates}

\author{
  Nils Matteson\\
  thaw\\
  \texttt{nils@thaw.sh}
}

\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
Per-token counterfactual credit estimation is the standard tool for asking
which token in a language-model rollout caused the final answer to be right
or wrong: cut the transcript at a pivot token, substitute an alternative,
replay continuations, and compare outcomes. Every published method we could
verify implements the replay by \emph{re-feeding} the transcript prefix
through the model as a fresh prompt, implicitly assuming that a re-fed prefix
reproduces the state the model actually passed through during generation. We
test that assumption directly on a stock inference engine, using a three-pass
design: continuations resumed from the exact decode-time KV state, an
identical second exact pass that serves as a replica noise floor, and a
re-feed pass in which the only changed variable is the prefill path. Across
four configurations and two models, including a GRPO-trained checkpoint,
re-feeding flips the sign of the credit estimate at decision tokens at rates
20--27 percentage points above the replica floor (paired
$z = 2.6$--$9.3$). At the probe level, resuming the decode-time state
reproduces the next-token logprob bit-exactly at 78\% of pivots; re-feeding
does so at 3.9\%. Even the floor is informative: batch-composition numerics
alone flip greedy outcomes at roughly 14\% of low-margin pivots. The
experiment was pre-registered with a locked decision rule and cost under
\$2 of compute. We recommend that counterfactual credit studies resume exact
decoder state (or use batch-invariant kernels) and report a replica floor;
without one, a substantial fraction of reported token-level credit signs are
attributable to the measurement procedure rather than the model.
\end{abstract}

\section{Introduction}

Reinforcement learning with verifiable rewards has made per-token credit
assignment a central measurement problem for language-model reasoning. Group
relative policy optimization \citep{shao2024deepseekmath} and its
descendants train on sequence-level rewards, so a growing line of work asks
the finer-grained question: which individual tokens in a rollout actually
caused success or failure? The standard answer is counterfactual replay.
Choose a pivot token, force an alternative continuation token, sample $K$
completions under each arm, grade them, and read the per-token causal effect
off the difference \citep{ruan2025cft, khandoga2026beyond, chen2026exact}.

Every implementation of this recipe that we could verify shares one
mechanical step: the transcript prefix up to the pivot is \emph{re-fed}
through the model as a fresh prompt, and the continuations branch from the
resulting state. The implicit assumption is that re-feeding tokens
$x_{1..t}$ reconstructs the internal state the model was in when it
originally generated token $t$. It does not, exactly. Floating-point
reduction order inside attention and matrix-multiply kernels depends on
batch composition and on how the sequence is split between prefill and
decode, so the re-fed state differs from the original decode-time state in
the low-order bits \citep{he2025nondeterminism}. The working assumption in
the credit-assignment literature is that these differences wash out of the
final estimate.

This paper measures whether they do. The answer is no: at the low-margin
decision tokens that credit methods specifically target, the re-feed
shortcut flips the \emph{sign} of the estimated credit at rates far above
what measurement noise explains.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{A three-pass audit design.} We compare credit estimates from
  (i) continuations resumed from the exact decode-time KV state, (ii) an
  identical second exact pass, and (iii) a re-feed pass. Pass (ii) is the
  load-bearing control: it converts ``how often do estimates disagree?''
  into ``how much disagreement is attributable to the prefill path
  \emph{beyond} replica noise?'' The design runs on a stock engine (vLLM
  \citep{kwon2023vllm}) with no kernel modifications.
  \item \textbf{Quantification.} Pooled over two pre-registered cohorts (40
  GSM8K problems, 1{,}564 pivots), re-feeding flips the credit sign at
  33.0\% of eligible pivots against a 12.2\% replica floor (excess
  $+20.7$pp, paired McNemar $z = 9.33$). The effect persists with greedy
  continuations that eliminate sampling noise entirely ($+26.5$pp) and on a
  GRPO-trained checkpoint ($+26.3$pp).
  \item \textbf{A bit-level account.} Resuming the decode-time state
  reproduces the post-pivot next-token logprob \emph{exactly} (delta
  $= 0.0$) at 78.0\% of 3{,}127 probes; re-feeding does so at 3.9\%, with a
  90th-percentile logprob shift of $4.6\times10^{-2}$.
  \item \textbf{The floor as a finding.} Even between two bit-identical
  request batches, continuous-batching numerics flip greedy outcomes at
  $\sim$14\% of low-margin pivots, quantifying a hazard for any
  single-sample evaluation at decision tokens.
\end{enumerate}

The experiment was pre-registered before any GPU run: temperature, $K$,
grading, the pivot filter, and the paper-versus-null decision rule were
locked in advance, and the one amendment (a difficulty screen, after a smoke
run showed the model saturates easy problems) is dated and was blind to
drift direction. Total compute cost was approximately \$2 on rented A100s.
Code, the pre-registration, and every per-pivot record are public.

\section{Related Work}

\paragraph{Counterfactual token credit.} Critical-token fine-tuning
\citep{ruan2025cft} forces alternative tokens and regenerates continuations
to identify tokens that change outcomes. \citet{khandoga2026beyond}
construct causal per-token credit for policy optimization and observe
qualitatively that group-relative advantages can carry the wrong sign.
\citet{chen2026exact} show that exact counterfactuals beat approximate ones
for credit assignment in cooperative LLM-agent systems, at the level of
whole agent actions. All of these estimate counterfactual state by re-feeding
or regenerating from text; none resume the original decoder state, and none
report a replica control that would bound the contribution of the replay
procedure itself. Our result does not diminish these methods; it quantifies
a systematic error term that all of them currently absorb silently.

\paragraph{Inference nondeterminism and training mismatch.}
\citet{he2025nondeterminism} identify batch invariance as the root cause of
LLM inference nondeterminism and fix it with batch-invariant kernels,
demonstrating bitwise-deterministic sampling and true on-policy RL. Their
victim is the serving endpoint and the trainer-versus-sampler gap in the RL
loop. \citet{zhong2026tim} diagnose training-inference mismatch, where the
rollout engine and the trainer assign different logprobs to the same tokens
under the same weights, as an independent cause of RL training collapse.
Both lines concern the \emph{training pipeline}. Our victim is different:
the \emph{measurement methodology} of the credit-assignment literature,
where the comparison pair is not trainer-versus-sampler but re-fed prefill
state versus original decode-time state, on the stock engines that
methodology actually runs on. The three lines compose: batch-invariant
kernels are the fix, mismatch diagnosis covers training, and this paper
covers measurement. In a fully batch-invariant engine where a token's
reduction order is independent of its prefill/decode split, the gap we
measure should collapse to the replica floor by construction; that is a
testable prediction of \citet{he2025nondeterminism}'s kernels which our
harness can evaluate directly.

\section{Method}

\subsection{Setup and estimator}

For each problem we sample $N = 8$ trunk rollouts at temperature $0.7$ with
top-$k$ logprobs recorded. Because the engine caches KV blocks as it
decodes, the exact state every rollout passed through remains resident in
the prefix cache at block granularity. A \emph{pivot} is a generated token
where the top-1/top-2 logprob gap is below $1.0$ (a low-margin decision
token, the population credit methods target), restricted to positions within
2 tokens past a 16-token block boundary so that the exact arm resumes from
cached state with at most 2 recomputed tokens. Per rollout we take at most
5 pivots, earliest first.

At pivot position $t$ with sampled token $a$ and best alternative $a'$, the
credit estimate under a given replay method is
\[
\At \;=\; \frac{1}{K}\sum_{k=1}^{K} R\big(x_{1..t-1},\,a,\,c_k\big)
\;-\;
\frac{1}{K}\sum_{k=1}^{K} R\big(x_{1..t-1},\,a',\,c'_k\big),
\]
where $c_k, c'_k$ are sampled continuations and $R$ grades the completed
solution by exact match on the GSM8K \citep{cobbe2021gsm8k} final answer.
Continuation seeds are a deterministic hash of (problem, rollout, pivot,
arm, $k$), so all replay methods receive identical sampling randomness; the
only uncontrolled difference between methods is the prefill path.

\subsection{Three passes}

\begin{description}
  \item[Exact pass 1.] All continuation requests for a problem are submitted
  while the trunk's decode-time KV blocks are resident. The prefix
  cache-hits the state the rollout actually traversed; continuations resume
  from it.
  \item[Exact pass 2 (replica floor).] The identical request batch is
  submitted again, cache still warm. Any disagreement between passes 1 and 2
  is attributable to continuous-batching numerics and sampling-on-perturbed-
  logits alone: it is the noise floor of the measurement itself.
  \item[Re-feed pass.] The prefix cache is reset and the identical batch is
  submitted a third time. The trunk is now freshly prefilled, exactly as in
  published credit pipelines. Differences beyond the replica floor are
  attributable to the re-feed prefill path.
\end{description}

The primary metric is the fraction of eligible pivots (those where at least
one pass yields $\At \neq 0$) at which $\mathrm{sign}(\At)$ under re-feed
disagrees with exact pass 1, reported against the same quantity for exact
pass 2. Each pass also issues one greedy 32-token probe per arm with top-2
logprobs recorded, giving a grading-free channel: probe-sequence divergence
rates and the absolute delta of the first post-pivot token's logprob.

\subsection{Why decode-time resume is the right reference}

The decode-time KV state is not an alternative approximation; it is the
state the model was actually in when it chose token $t$. A counterfactual
\emph{do}-operation at $t$ is defined relative to that state. Resuming it
requires no kernel modifications: paged-attention engines already retain it
in the prefix cache, and durable snapshot tools can persist it across
processes. The replica pass empirically validates the reference: 78\% of
resumed probes reproduce the next-token logprob bit-exactly
(Section~\ref{sec:results}), whereas a re-fed prefix essentially never does.
One honest caveat: blocks are 16 tokens, so ``exact'' means exact to the
block boundary, with at most 2 trailing tokens recomputed under our pivot
filter.

\subsection{Pre-registration}

Temperature, $K$, the grading function, the pivot filter, and the decision
rule (sign-flip excess above 5\%: the method is materially unsound; below
2\%: re-feed is adequate) were locked before any GPU run. A smoke run
revealed that Qwen2.5-7B-Instruct solves easy GSM8K problems in every
continuation, making $\At = 0$ everywhere and the metric undecidable for
lack of eligible pivots. Amendment 1, dated before any full-run results,
added a second cohort screened only on \emph{trunk} rollout accuracy being
mixed (neither $0/8$ nor $8/8$), a criterion computed before any arm
continuations exist and blind to drift direction. Both cohorts are reported
separately and pooled.

\section{Experimental Setup}

\textbf{Models.} Qwen2.5-7B-Instruct \citep{qwen2025} and
Qwen-2.5-7B-SimpleRL-Zoo \citep{zeng2025simplerlzoo}, a GRPO-trained
checkpoint built from the Qwen2.5-7B base on mathematical reasoning.
\textbf{Engine.} Stock vLLM 0.22.1 \citep{kwon2023vllm}, bfloat16, prefix
caching enabled, no kernel modifications. \textbf{Hardware.} One A100 SXM
80GB per run (rented; two pods, roughly 80 minutes total, $\sim$\$2).
\textbf{Data.} GSM8K test split \citep{cobbe2021gsm8k}: the first 20
problems (primary cohort) and the first 20 mixed-accuracy problems under the
Amendment 1 screen (screened cohort). \textbf{Configurations.} Main:
temperature 0.7, $K=4$. Secondary: greedy continuations (temperature 0,
$K=1$), $K=8$, and the GRPO checkpoint, all on screened cohorts.

\section{Results}
\label{sec:results}

\subsection{Primary metric: sign flips above the replica floor}

Table~\ref{tab:main} reports, on the common eligible set of each run (pivots
where any pass gives $\At \neq 0$, identical denominator for both
comparisons), the rate at which re-feed disagrees in sign with exact pass 1,
against the exact-versus-exact replica floor, with a paired McNemar test on
the discordant pivots.

\begin{table}[t]
\centering
\small
\begin{tabular}{llrrrrr}
\toprule
Configuration & Model & $n$ & Re-feed flip & Floor & Excess & $z$ \\
\midrule
Main, pooled (t=0.7, $K{=}4$) & Qwen2.5-7B-Instruct & 622 & 33.0\% & 12.2\% & $+20.7$pp & 9.33 \\
Greedy (t=0, $K{=}1$)         & Qwen2.5-7B-Instruct & 189 & 40.2\% & 13.8\% & $+26.5$pp & 5.66 \\
$K{=}8$ (t=0.7)$^{\dagger}$   & Qwen2.5-7B-Instruct & 104 & 35.6\% & 22.1\% & $+13.5$pp & 2.56 \\
GRPO ckpt (t=0.7, $K{=}4$)    & SimpleRL-Zoo-7B     & 373 & 35.4\% & 9.1\%  & $+26.3$pp & 8.66 \\
\bottomrule
\end{tabular}
\caption{Sign-flip rate of the credit estimate under re-feed versus the
exact-replica noise floor, on each run's common eligible set. $z$ is the
paired McNemar statistic over discordant pivots. $^{\dagger}$Truncated at
4/20 problems by a transient pod disk failure unrelated to the harness;
reported for completeness.}
\label{tab:main}
\end{table}

Every configuration is positive and significant. The two strongest are the
two that answer the obvious objections. With greedy continuations there is
no sampling randomness at all, and re-feed still flips 40.2\% of estimates
against a 13.8\% pure-numerics floor; per problem, 11 of 19 problems show
re-feed worse than floor and zero show the reverse. On the GRPO-trained
checkpoint, the population for which credit assignment is actually studied,
the excess is $+26.3$pp and the per-problem split is 16 worse, 2 better.
In the pooled main configuration the discordant counts are 160 pivots
flipped only by re-feed against 31 flipped only by the replica
($z = 9.33$). The effect is flat in pivot margin: 32.6\% below a 0.3
logprob gap versus 33.2\% above it (floor 12.2\% in both buckets), so this
is not confined to near-ties.

\subsection{Bit-level account}

The greedy probes give a grading-free view of the same phenomenon. Across
3{,}127 probes pooled over the full runs, the replica pass reproduces the
first post-pivot token's logprob \emph{exactly} (absolute delta $= 0.0$) at
78.0\% of pivots; the re-feed pass does so at 3.9\%. When nonzero, the
median replica delta and median re-feed delta are both
$\sim 1.7\times10^{-3}$, but the tails differ by an order of magnitude: the
90th-percentile delta is $2.5\times10^{-3}$ for the replica versus
$4.6\times10^{-2}$ for re-feed (maximum $0.18$). Probe sequences diverge
within 32 greedy tokens at 23.1\% of pivots under re-feed versus 6.7\% for
the replica on the main runs, and at 16.0\% versus 1.7\% (a $9.4\times$
ratio) on the GRPO checkpoint.

\subsection{The floor is also a finding}

Two bit-identical request batches, submitted seconds apart against the same
warm cache with the same seeds, disagree on the sign of $\At$ at 9--14\% of
eligible pivots (22\% under $K{=}8$'s partial run) and on greedy
continuations at 6--9\% of pivots. This is \citet{he2025nondeterminism}'s
batch-composition nondeterminism expressed in the unit that matters for
evaluation: at the decision tokens where models are uncertain, batching
numerics alone flip single-sample outcomes at a rate that would be
catastrophic to interpret as signal. Any study that evaluates at
low-margin tokens without a replica control is reading a coin flip at
roughly one pivot in ten.

\section{Discussion}

\paragraph{What this does and does not show.} It shows that the re-feed
shortcut is materially unsound as a measurement procedure for per-token
counterfactual credit at decision tokens on stock engines: roughly one in
five sign judgments beyond noise is an artifact of the replay path. It does
not show that GRPO training is broken (we measured the estimator, not
training outcomes), and it does not claim novelty for the existence of
batch-composition drift, which \citet{he2025nondeterminism} established and
fixed at the kernel level. The contribution is the quantification, on the
measurement methodology the credit literature actually uses, with a control
that no prior credit study reports.

\paragraph{Recommendations.} (1) Resume the decode-time state for
counterfactual replay. On paged-attention engines this is engine-native via
the prefix cache within a process, and durable session-snapshot tooling
extends it across processes. (2) Alternatively, run on batch-invariant
kernels \citep{he2025nondeterminism}, under which prefill/decode-split
invariance should collapse the re-feed gap to the floor; confirming that
prediction with this harness is immediate future work. (3) In all cases,
report an exact-replica floor. A credit-assignment result at low-margin
tokens without a floor is uninterpretable, since the floor alone is
$\sim$10\%. (4) Pre-register estimator settings: the flip rate is sensitive
to temperature, $K$, grading, and the pivot filter, and is easy to tune
post hoc.

\paragraph{Limitations.} One task family (GSM8K), one engine version (vLLM
0.22.1), one GPU type (A100); pivots are deliberately conditioned on low
margin and on block-boundary proximity, so the rates are statements about
decision tokens, not about all tokens; $K$ is small (the binary-graded
estimator is noisy, which the floor makes explicit); the $K{=}8$ run is a
4-problem partial; and ``exact'' is exact to the 16-token block boundary.
None of these caveats favor the effect: the strongest results occur exactly
where the caveats are weakest (greedy continuations, the GRPO checkpoint).

\section{Reproducibility}

The pre-registration (with its dated amendment), the harness (a single
$\sim$500-line Python file against stock vLLM), every per-pivot record of
all five runs, and the run logs are public in the thaw repository
(\url{https://github.com/thaw-ai/thaw}, \texttt{benchmarks/}). Total
compute: two rented A100 pods, roughly 80 minutes and \$2. Each headline
number in this paper can be recomputed from the published JSONs with the
included analysis function.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
